Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2014-09-29 article.Rmd
Deleted faceted chart section for AHR

The path from a dissertation to a book review in the *American Historical Association* is not automatic. Not all dissertations are published as books, not all books are submitted to the *American Historical Review*, and not all book submissions are accepted by the *AHR* editors for review. As of September 2014 the journal's website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)." [American Historical Review n.d.] Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2010 survey by the American Historical Association revealed the female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. More troublingly, women continue to face gender discrimination in the professional sphere.

@@

The *AHR* also categorizes its reviews under different historical sub-fields. Although the journal’s taxonomy has changed quite a bit over time, it did allow us to refine our analysis. Were there more reviews of female authors in certain sub-fields rather than others, and has that changed over the past forty years? Caribbean/Latin American history has had something approaching equal representation for the past decade-and-a-half. In both African history and Ancient/Medieval history female historians made some quite dramatic gains during the late-nineties and aughts. The most uneven parties, however, are also the two subject categories that publish the most book reviews in the *AHR*: Modern/Early Modern Europe and the United States/Canada. Both sub-fields have made progress over the four decades but still hover at around two-thirds male. Gender representation is not evenly distributed across the profession.

```{r reviews-by-category}
categories <- read.csv("data/ahr-categories.csv", stringsAsFactors = FALSE)

ahr_predicted %>%
  left_join(categories, by = "category") %>%
  filter(!our_category %in% c(NA, 
                              "Articles",
                              "Latin and South America, Carribean",
                              "Middle East")) %>%
  select(year, gender_author, our_category) %>% 
  group_by(year, gender_author, our_category) %>%  
  summarize(n = n()) %>%
  spread(gender_author, n, fill = 0) %>% 
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion,
         year, our_category) %>% 
  gather(gender, proportion, -year, -our_category) %>%
ggplot(aes(x = year, y = proportion, fill = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  facet_wrap(~ our_category) +
  xlim(1970, 2012) +
  ggtitle("Gender of AHR authors by book category, 1970-2012") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_brewer(palette = article_palette,
                    labels = c("Female", "Male", "Unknown"),
                    name = "Gender") +
  article_theme +
  theme(strip.background = element_rect(color = "white"))
```

>>>>

The path from a dissertation to a monograph to a book review in the *American Historical Association* isn't automatic. Not all dissertations are published as books, not all books are submitted to the *American Historical Review*, and not all book submissions are accepted by the *AHR* editors for review. As of September 2014 the journal's website reminds readers that "the AHR receives over 3,000 books a year; we have the resources to publish at most 1,000 reviews a year (approximately 200 per issue)." [American Historical Review n.d.] Fewer and fewer female historians make it through each successive step. Establishing a single causal reason for this pattern is all but impossible, and instead a murky concoction of interrelated factors likely contributes to the gap. For one, a 2010 survey by the American Historical Association revealed the female historians dedicated substantially more time to child and elder care than their male colleagues, leaving less time for research. This gender gap is more likely to impact the production of a book than a dissertation, as graduate students are generally younger and less likely to be juggling child-rearing responsibilities while writing than their older peers. More troublingly, women continue to face gender discrimination in the professional sphere.

@@
